Applied Measure Theory for Probabilistic Modeling

### 2022-05-11 paper/paper.tex
edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Markov chains}

Some kernels $\kappa$ map elements of a state space $X$ to measures on the same space.

In this case, repeated application of the kernel ``walks'' through $X$, yielding the law of a \href{https://en.wikipedia.org/wiki/Markov_chain}{\emph{Markov chain}}. We support this in \verb|MeasureTheory| through the \verb|Chain| combinator. This takes a kernel (or function) and a starting measure, and can be used with Julia's convenient \verb|do| notation. For example, we can write a Gaussian random walk as

\begin{verbatim}
    mc = Chain(Normal(μ=0.0)) do x Normal(μ=x) end
\end{verbatim}

A random sample from a \verb|Chain| is an infinite sequence expressed as a iterator, e.g.

\begin{verbatim}
    julia> r = rand(mc);

    julia> x = Iterators.take(r, 3) |> collect
    3-element Vector{Any}:
     -0.4931543737034523
     -0.5661895116186417
     -1.3286977670590228
\end{verbatim}

Thus, for example, we can evaluate \verb|mc| on the \verb|x| we just computed:
\begin{verbatim}
    julia> logdensity(mc, x)
    -0.4149771036439342
\end{verbatim}

Crucially, the call to \verb|rand| builds a struct containing a random seed, to ensure that repeated realizations of \verb|r| will be identical.

To give us more control over the sampling procedure, \verb|Chain| is implemented using \emph{dynamic iterators} from \verb|DynamicIterators.jl| \cite{DynamicIterators.jl}.

The appropriate $\sigma$-algebra for infinite sequences requires some care. A Markov chain on $X$ has $\sigma$-algebra generated by sets of \emph{sequences with finitely many constraints}.\footnote{That is,  the smallest $\sigma$-algebra containing all sets of sequences which can be written as $\{x \colon x_i \in A_i \mid i \in I\}$ using collections of measurable sets $\{A_i \mid i\in I\}$ indexed by finite index sets $I$, called \href{https://en.wikipedia.org/wiki/Cylinder_set_measure}{\emph{cylinder sets.}}}

@@

A special case of this is when $\mu = \nu$. In \verb|MeasureTheory| we refer to this as a \emph{power measure}, written \verb|μ ^ 2| or \verb|μ ^ n| for higher dimensions. More generally, the second argument can be a \verb|Tuple|, so for example \verb|μ ^ (2, 3)| extends $\mu$ as a product measure over $2\times 3$ matrices.

@@

Narayanan et al \cite{narayanan2016probabilistic} describe \verb|Hakaru|, a system for Bayesian modeling using measures. Here, a measure is a functional
\[
\mu[f] = \int f\ \dd\mu\ ,
\]
represented as a program. Hakaru's combinators are then expressed as compilers taking programs as the inputs.

Radul and Alexeev \cite{Radul2020} describe the \emph{base measure problem} of losing track of a base measure when applying a transformation, and suggest standardizing around Hausdorff measure as a solution. This problem doesn't arise for us, because the base measure is always taken into account.

Borgstr\"om et al \cite{Borgstr_m_2013} describes the \verb|Fun| system in F\# in terms of \emph{measure transformer semantics}, but discusses only \emph{finite} measures.

@@

In Julia \cite{bezanson2017julia}, the \verb|Distributions.jl| package \cite{Distributions.jl-2019} is very popular for computations on distributions. The drawbacks of \verb|Distributions.jl| are essentially those described in the first few sections of this paper. 
Current advantages over \verb|MeasureTheory| are the extensive range of distributions it implements and its popularity and familiarity to many Julia users.

\verb|MeasureTheory.jl| currently has \verb|Distributions.jl| as a dependency, and uses it as a fall-back for many computations. For convenience, we also re-export the \verb|Distributions| module under the handle \verb|Dists|.

>>>>

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Markov chains}

% Some kernels $\kappa$ map elements of a state space $X$ to measures on the same space.

% In this case, repeated application of the kernel ``walks'' through $X$, yielding the law of a \href{https://en.wikipedia.org/wiki/Markov_chain}{\emph{Markov chain}}. We support this in \verb|MeasureTheory| through the \verb|Chain| combinator. This takes a kernel (or function) and a starting measure, and can be used with Julia's convenient \verb|do| notation. For example, we can write a Gaussian random walk as

% \begin{verbatim}
%     mc = Chain(Normal(μ=0.0)) do x Normal(μ=x) end
% \end{verbatim}

% A random sample from a \verb|Chain| is an infinite sequence expressed as a iterator, e.g.

% \begin{verbatim}
%     julia> r = rand(mc);

%     julia> x = Iterators.take(r, 3) |> collect
%     3-element Vector{Any}:
%      -0.4931543737034523
%      -0.5661895116186417
%      -1.3286977670590228
% \end{verbatim}

% Thus, for example, we can evaluate \verb|mc| on the \verb|x| we just computed:
% \begin{verbatim}
%     julia> logdensity(mc, x)
%     -0.4149771036439342
% \end{verbatim}

% Crucially, the call to \verb|rand| builds a struct containing a random seed, to ensure that repeated realizations of \verb|r| will be identical.

% To give us more control over the sampling procedure, \verb|Chain| is implemented using \emph{dynamic iterators} from \verb|DynamicIterators.jl| \cite{DynamicIterators.jl}.

% The appropriate $\sigma$-algebra for infinite sequences requires some care. A Markov chain on $X$ has $\sigma$-algebra generated by sets of \emph{sequences with finitely many constraints}.\footnote{That is,  the smallest $\sigma$-algebra containing all sets of sequences which can be written as $\{x \colon x_i \in A_i \mid i \in I\}$ using collections of measurable sets $\{A_i \mid i\in I\}$ indexed by finite index sets $I$, called \href{https://en.wikipedia.org/wiki/Cylinder_set_measure}{\emph{cylinder sets.}}}

@@

A special case of this is when $\mu = \nu$. In \verb|MeasureTheory| we refer to this as a \emph{power measure}, written \verb|μ^2| or \verb|μ^n| for higher dimensions. More generally, the second argument can be a \verb|Tuple|, so for example \verb|μ^(2,3)| extends $\mu$ as a product measure over $2\times 3$ matrices.

@@

While the vast majority of research in computational statistics works explicitly in terms of probability distributions, a few authors have addressed measures more generally:

\begin{itemize}
\item Borgstr\"om et al \cite{Borgstr_m_2013} describes the \verb|Fun| system in F\# in terms of \emph{measure transformer semantics}, but discusses only \emph{finite} measures.

\item Narayanan et al \cite{narayanan2016probabilistic} describe \verb|Hakaru|, a system for Bayesian modeling using measures. Here, a measure is a functional
\[
\mu[f] = \int f\ \dd\mu\ ,
\]
represented as a program. Hakaru's combinators are then expressed as compilers taking programs as the inputs.

\item Radul and Alexeev \cite{Radul2020} describe the \emph{base measure problem} of losing track of a base measure when applying a transformation, and suggest standardizing around Hausdorff measure as a solution. This problem doesn't arise for us, because the base measure is always taken into account.

\end{itemize}

In Julia \cite{bezanson2017julia}, the \verb|Distributions.jl| package \cite{Distributions.jl-2019} is very popular for computations on distributions. The drawbacks of \verb|Distributions.jl| are essentially those described in the first few sections of this paper. 
Current advantages over \verb|MeasureTheory| are the extensive range of distributions it implements and its popularity and familiarity to many Julia users. \verb|MeasureTheory.jl| currently has \verb|Distributions.jl| as a dependency, and uses it as a fall-back for many computations. For convenience, we also re-export the \verb|Distributions| module under the handle \verb|Dists|.

@@
